---
title: "National Risk Index and ACS in California"
format: 
  html:
    theme: darkly
author: 'Peter Vitale'
date: '2026-02-11'
---
## Load Libraries
```{r}
pacman::p_load('here',
               'janitor',
               'tidyverse',
               'ggthemes',
               'ggbeeswarm',
               'patchwork',
               'gghighlight',
               'ggridges')
```

## Read Data with api key 
```{r}
#| warning: false
#....Step 1a: see all available ACS variables + descriptions.....
#acs_vars <- tidycensus::load_variables(year = 2023,
#                                       dataset = "acs1")

#..............Step 1b: import race & ethnicity data.............
#race_ethnicity <- tidycensus::get_acs(
#  geography = "county",
#  survey = "acs1",
  # NOTE: you may not end up using all these variables
#  variables = c("B01003_001", "B02001_002", "B02001_003", 
#                "B02001_004", "B02001_005", "B02001_006",
#                "B02001_007", "B02001_008", "B03002_012",
#                "B03002_002"),
#  state = "CA", 
#  year = 2023) |>
  # join variable descriptions (so we know what's what!)
#  dplyr::left_join(acs_vars, by = dplyr::join_by(variable == name)) 

#.................Step 2: write ACS data to file.................
# readr::write_csv(race_ethnicity, here::here("data", "ACS-1yr-2023-county-race-ethnicity.csv"))

#..................Step 3: read in your CSV file.................
race_ethnicity <- readr::read_csv(here::here("data", "ACS-1yr-2023-county-race-ethnicity.csv")) %>% 
  clean_names() %>% 
  rename(nri_id = geoid)
```
## Read NRI csv
```{r} 
#| warning: FALSE
nri_df <- read_csv(here('data', 
                        'National_Risk_Index_Counties_807384124455672111.csv')) %>% 
  clean_names()
```
## Clean NRI Data

```{r}
nri_clean <- nri_df %>% 
  select(c(1,2,3,4,6,7,10, 15,16)) %>% 
  clean_names() %>% 
  filter(state_name == 'California') %>% # Keep only California 
  separate(national_risk_index_id, 
           into = c(NA, 'nri_id'),
           sep =  'C') 
```

## Join NRI and race_ethnicity
```{r}
nri_acs_joined <- left_join(nri_clean, 
                            race_ethnicity,
                            by = 'nri_id') %>% 
  filter(!is.na(name))
```

Create a data viz that helps to answer the question, How does climate hazard risk exposure vary across racial / ethnic groups in California? 

```{r}
nri_county_race <- nri_acs_joined %>% 
   mutate(label = case_when( # Mutate the groupings to more easy to use labels
    label == "Estimate!!Total" ~ 'Overall Population',
    label == "Estimate!!Total:!!White alone"  ~ 'White',
    label == "Estimate!!Total:!!Black or African American alone" ~ 'Black',
    label == "Estimate!!Total:!!American Indian and Alaska Native alone" ~ 'Native',
    label == "Estimate!!Total:!!Asian alone" ~ 'Asian',
    label == "Estimate!!Total:!!Native Hawaiian and Other Pacific Islander alone" ~ 'Pacific Islander',
    label == "Estimate!!Total:!!Two or More Races:" ~ 'Multi-Ethnic',
    label == "Estimate!!Total:!!Hispanic or Latino:" ~ 'Hispanic',
    TRUE ~ 'Other Race'
  ))  %>%
  # Remove where those labels are of the entire population in a county
  filter(label != 'Overall Population') %>% 
  mutate(nri_rating = factor(national_risk_index_rating_composite,
                                levels = c('Very High',
                                           'Relatively High',
                                           'Relatively Moderate',
                                           'Relatively Low')))
  
  
```

## Make the plot

```{r}
#| warning: FALSE
#| fig-width: 12
#| fig-height: 7
#| fig-align: "center"
#| fig-alt: 'A histogram of the National Risk Index (NRI) composite scores compared to American Community Survey(ACS) defined racial and ethnic groups. The plot has NRI on the x-axis ranging from 80 to 100. The y-axis contains the proportional value of the populations in each racial/ethnic group. The bins of the histogram are 1 point of NRI. A majority of the NRI scores are close to 100. The histogram shows no pattern in racial distribution over increasing NRI scores.'

showtext::showtext_auto(enable = TRUE)
ggplot(nri_county_race,
       aes(y = national_risk_index_score_composite, 
           x = label, 
           width= .7, 
           fill = nri_rating))+
  geom_histogram(
                 stat="identity")+
  scale_fill_manual(values = c("#448aff", # Scale the color with coolors color scale
                               "#1565c0",
                               "#009688",
                               "#8bc34a"))+
  # Update the labels
  labs(y = 'Cumulutive National Risk Index Over All Counties',
       fill = 'Risk Index Rating',
       title = 'Composite National Risk Index Score In California',
       subtitle = 'Non-Identified Races show higher NRI Scores, with high proportions of Relatively High \ncompared to other racial and ethnic groups',
       caption = 'Data: FEMA National Risk Index (2025 Release) \nand American Community Survey Data (2023 Release)' )+
  theme_tufte()+
  # I want dark colors so I change the background to black and the text to white
  theme(plot.background = element_rect(fill = "#222222", 
                                       color = "#222222"),
        plot.title = element_text(color = 'white', 
                                   size = 26,
                                  face = 'bold',
                                  family = 'serif'), # Times new roman 
        plot.subtitle = element_text(color = 'white',
                                     size = 20,
                                     face = 'italic',
                                     family = 'serif'),
        axis.title = element_text(color = 'white', 
                                   size = 18,
                                  family = 'serif'),
        legend.title = element_text(color = 'white',
                                     size = 18,
                                    family = 'serif'),
        plot.caption =  element_text(color = 'white',
                                     size = 12,
                                     hjust = 1.44),
        axis.title.y = element_text(margin = margin(r = 15, 
                                                    unit = "pt")),# Add margin on x
        axis.title.x = element_blank(),
        axis.text = element_text(color = 'white',
                                 size = 16),
        axis.text.x = element_text(angle = 45,
                                   vjust = .6),
        legend.text = element_text(color = 'white',
                                   size = 16))
```
1. What are your variables of interest and what kinds of data (e.g. numeric, categorical, ordered, etc.) are they (a bullet point list is fine)?

I am interested in the `label`, `national_risk_index_score_composite`, and `estimate`

-  `label`: The categorical character race/ethnicity label provided by the ACS

-  `national_risk_index_score_composite`: Numeric county specific index score.

-  `estimate`: Numeric estimate of population in each `label`

2. How did you decide which type of graphic form was best suited for answering the question? What alternative graphic forms could you have used instead? Why did you settle on this particular graphic form?

I played around a couple of ideas. My first thought was a streamgraph, however the inputs to that need a date. I could have also gone in the donut direction but I had a vision of that streamgraph aesthetic. 

3. Summarize your main finding in no more than two sentences.

There is no discernable trend in population makeup with increasing NRI score. 

4. What modifications did you make to this visualization to make it more easily readable?

I changed the font family, as well as making the plot and QMD a dark background. I did this because I wanted to use a bright color palette for the histogram, but when rendered it was almost blinding. Dark backgrounds often help me in terms of readability, so I hope the effect is the same to others. 

5. Is there anything you wanted to implement, but didn’t know how? If so, please describe.

Like I mentioned, I wasn't able to make a streamgraph due to not having a date. I tried to coerce the counties into dates (like days of the year) but that was not a scalable solution. Ultimately I shifted gears, but I am counting down the days until I can run a streamgraph. 



